---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mesoclim
<!-- badges: start -->
<!-- badges: end -->

## Installation

You can install the development version of mesoclim from [GitHub](https://github.com/) with:

``` {r install }
# install.packages("devtools")
# devtools::install_github("ilyamaclean/mesoclim")
```

## Overview
The mesoclim package provides functions to enable the mechanistic downscaling of climate data from coarse temporal and/or spatial resolutions to finer resolutions at which mesoclimatic processes, such as the effects of elevation, cold air drainage and coastal exposure, significantly affect how local weather conditions vary across the landscape. 

The package permits the downscaling of climate data with a spatial resolution of 10s of kilometers to resolutions of under a single kilometer. It does not permit the modelling of microclimatic processes or below-canopy or below-ground conditions, but provides suitable data for subsequent microclimate modelling (refs to other packages).

The functions are organised around key steps in the downscaling work flow:

### 1 Downloading climate data inputs
Functions exist to download common sources of historic and future global climate and ancillary data from available repositories. Typically this data is have spatial resolutions of 10s of kilometers and a daily or hourly temporal resolution. 

See: `vignette("mesoclim_1_download")`

### 2 Preprocessing data inputs
Functions generate standard inputs from source files of coarse resolution climate and ancillary data. These inputs are required for the subsequent mesoclimate downscaling. Functions are also provided to check inputs and provide tabular and graphical summaries of the input data.
{_Something on using weather station inputs??_}

See: `vignette("mesoclim_2_preparedata")`

### 3 Spatial downscaling
Functions capture the effect of specific processes such as elevation, coastal exposure or cold air drainage, whereas wrapper functions apply an ensemble of chosen processes to downscale climate data to finer resolutions. Available function allow a high degree of control of which mesoclimatic effects are used in downscaling, reflecting user interest and/or whether particular processes have been partially captured in the input datasets. 

See: `vignette("mesoclim_3_spdscale")`

### 4 Temporal downscaling
Functions allow the conversion of daily to hourly estimates of climate conditions.

See: `vignette("mesoclim_4_tmpdscale")`

### 5 Bias correction
Functions are provided to statistically compare different climate datasets and correct one set of data against another. Typically these functions will be used to correct modelled data using a long timeseries of observational data. 

See: `vignette("mesoclim_5_biascorrect")`

Functions are also provided to carry out simple data checking, statistical summaries and graphing of climate datasets.




