---
title: "3. Spatial downscaling"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup, include=FALSE}
#library(mesoclim)

devtools::load_all(quiet=TRUE)
```

# 3. Spatial Downscaling of UKCP18 regional climate data

## Defining the area of interest
The projection, resolution and extent of the downscaled area is defined by providing a fine-resolution DTM that also acts as a land/sea mask with sea cells having a NA value.


```{r downscaled-dtm,  results="hide"}

# Get fine and medium res DTMs from package
dir_datasets<-system.file('data',package='mesoclim')

load(file.path(dir_datasets,'dtmf.rda'))
load(file.path(dir_datasets,'dtmm.rda'))

dtmm<-unwrap(dtmm)
dtmf<-unwrap(dtmf)
```

## Preparing the data
Coarse resolution climate and ancillary data will need to be prepared as outlined in `vignette('mesoclim_2_preparedata', package='mesoclim')`.  

Sea surface temperature data must also be prepared for the surrounding area and for the same time period as the input climate data. 

```{r sst-prepare}
startdate<-as.POSIXlt('2018/01/01')
enddate<-as.POSIXlt('2018/12/31')
member<-'01'

#Load prepared climate data
ukcp18rcm<-read_climdata(file.path(system.file('extdata/preprepdata',package='mesoclim'),'ukcp18rcm.Rds'))

# Load sea surface data 
dir_sst<-system.file('extdata/ukcp18sst',package='mesoclim')
ukcp18sst<-create_ukcpsst_data(dir_sst,startdate,enddate,ukcp18rcm$dtmc,member)

# Interpolate SST data to required timesteps
sst<-spatial_interpNA(ukcp18sst)
sst<-time_interp(sst,ukcp18rcm$tme)
```

## Single step downscaling
Wrapper function allows downscaling in a single function call, with parameters defining which processes are captured by the downscaling.

### Improvements
* Include dtmf in `spatialdownscale` output.

* Allow `spatialdownscale` output as spatrasters, packedspatrasters or arrays.

* Include wind and temp height in `spatialdownscale` output

* Write climdata to file function that converts to packed rast or arrays?

* Read climdata file function?

```{r onestep-spdown}

# Downscale one year of future climate data

t0<-now()
mesolizd<-spatialdownscale(ukcp18rcm, sst, dtmf, dtmm, basins = NA, cad = TRUE,
                           coastal = TRUE, refhgt = ukcp18rcm$tempheight_m, 
                           uhgt = ukcp18rcm$windheight_m, 
                           rhmin = 20, pksealevel = TRUE, patchsim = TRUE, 
                           terrainshade = FALSE, 
                           precipmethod = "Tps", fast = TRUE, noraincut = 0.01)
print(now()-t0)

# write_climdata(mesolizd,file.path(system.file('extdata/dscaledata',package='mesoclim'),'mesolizd.rds'))

# mesolizd<-read_climdata(file.path(system.file('extdata/dscaledata',package='mesoclim'),'mesolizd.rds')))
```

### Display outputs

In tabular form for all times and locations:

```{r output_table}
smry_fun<-function(x) summary(as.vector(.is(x)))
rslt<-sapply(mesolizd[names(mesolizd)],smry_fun)
stats_df<-as.data.frame(t(round(rslt,3)))[,c('Min.','Mean','Max.')]
print(stats_df)
  
```

Or as selected rasters for days corresponding to spatial quantiles (eg days where spatial means are ata min, median and max values):

__CHECK:__ Max diurnal temperature range plot shows grid effect?

```{r map-outputs}
for(var in names(mesolizd)){
  r<-terra::unwrap(mesolizd[[var]])
  names(r)<-rep(var,nlyr(r))
  plot_q_layers(r)
}

# Show spatial range in daily temperatures
plot(max(unwrap(mesolizd$tmax)-unwrap(mesolizd$tmin)),main='Max diurnal temperature range', font.main = 1, nc=1)

```

```{r tseries_output}
par(mar=c(1,1,1,1),cex.main=0.8, mgp=c(3,0.1,0))
layout(matrix(c(1,2,3,4,1,5,6,7,1,8,9,10),ncol=3),heights=c(1,3,3,3))

#par(mfrow=c(3,3))
plot.new()
text(0.5,0.5,"Spatial mean (green), max (red) and min (blue) by day of year",cex=1,font=1)

for(v in names(mesolizd)){
  r<-unwrap(mesolizd[[v]])
  plot_timestats_r(r,v,idx='doy',lgd=FALSE)
}

```


## Multiple step downscaling
Spatial downscaling can be decomposed to the effects of specific physical processes, such as the effect of elevation, coastal or cold air drainage on temperature. Specific functions for downscaling particular effects can be applied without needing to undertake full downscaling using all processes.

_Note:_ for individual functions, outputs may need further processing to ensure names are times are carried through.

### Improvements
* Allow option of spatrast or array outputs? Or enforce a standard spatrast output?

* Allow passing standard data input list of climdata as well as individual variables

* Allow variable winddir in windspeed downscaling - or warning message if input wind dir varies?


### Pressure 
```{r pressure_dscale}
pres<-presdownscale(ukcp18rcm$climarray$pres, dtmf, ukcp18rcm$dtmc, sealevel = TRUE)
names(pres)<-rep('pres',nlyr(pres))
terra::time(pres)<-ukcp18rcm$tme

plot_q_layers(pres)
```


### Cold air drainage 
Requires the estimation of drainage basins defined by topography which can be carried out separately from the estimation of cold-air drainage in downscaling.
```{r calc_basins}
# Basins - dtmf should have sea marked as NA
basins<-basindelin(dtmf,boundary=2)
plot(basins,main='basins')
```

For the chosen study area, cold air drainage has very little effect on local temperatures:
```{r calc_cad}
# Calculate using tmin to illustrate
ukcp18rcm$climarray$temp<-ukcp18rcm$climarray$tmin
tcad<-.tempcad(ukcp18rcm,dtmf,basins,refhgt=ukcp18rcm$tempheight_m)

names(tcad)<-rep('temp_cad',nlyr(tcad))
terra::time(tcad)<-ukcp18rcm$tme

plot_q_layers(tcad)
```

### Temperature elevation downscaling
Variable lapse rates derived from temperature, humidity and pressure are applied to correct for elevation effects when downscaling. For daily  temperatures this can be carried out for min and max daily values.
```{r temp_elev_dscale}
tminelev<-.tempelev(ukcp18rcm$climarray$tmin,dtmf,ukcp18rcm$dtmc,ukcp18rcm$climarray$relhum,ukcp18rcm$climarray$pres)
tmaxelev<-.tempelev(ukcp18rcm$climarray$tmax,dtmf,ukcp18rcm$dtmc,ukcp18rcm$climarray$relhum,ukcp18rcm$climarray$pres)

names(tminelev)<-rep('tmin_elev',nlyr(tminelev))
terra::time(tminelev)<-ukcp18rcm$tme

names(tmaxelev)<-rep('tmax_elev',nlyr(tmaxelev))
terra::time(tmaxelev)<-ukcp18rcm$tme

plot_q_layers(tminelev)
plot_q_layers(tmaxelev)

# Check dirunal range of elev downscaling
diurnaltmp<-tmaxelev-tminelev
names(diurnaltmp)<-rep('diurnal_t_range',nlyr(diurnaltmp))
plot_q_layers(diurnaltmp)

```

### Wind
Windspeed downscaling aims to capture the effect of both elevation and the sheltering effects of topography that are in turn dependent on wind direction and height above ground. 

```{r wind_dscale}
windspeed<-winddownscale(ukcp18rcm$climarray$windspeed, ukcp18rcm$climarray$winddir, dtmf, dtmm, ukcp18rcm$dtmc, uz = ukcp18rcm$windheight_m)

names(windspeed)<-rep('windspeed',nlyr(windspeed))
terra::time(windspeed)<-ukcp18rcm$tme

plot_q_layers(windspeed)

```

### Coastal effects
Uses sea surface temperatures, DTMs where sea is indicated by NA values and downscaled windspeed to estimate the effect of coastal effects.
Coarse, medium and fine resolution DTMs are used to determine effects of upwind sea areas.

_ADD:_ Guidance on selection of the extent of dtmm - how far should dtms extend from the downscaling area to adequately capture coastal effects??

```{r coastal}
sstf<-.resample(sst,dtmf)
tmincoast<-.tempcoastal(tminelev,sstf,windspeed,ukcp18rcm$climarray$winddir,dtmf,dtmm,ukcp18rcm$dtmc)
tmaxcoast<-.tempcoastal(tmaxelev,sstf,windspeed,ukcp18rcm$climarray$winddir,dtmf,dtmm,ukcp18rcm$dtmc)

names(tmincoast)<-rep('tmin_coastef',nlyr(tmincoast))
terra::time(tmincoast)<-ukcp18rcm$tme
names(tmaxcoast)<-rep('tmax_coastef',nlyr(tmaxcoast))
terra::time(tmaxcoast)<-ukcp18rcm$tme

plot_q_layers(tmincoast)
plot_q_layers(tmaxcoast)
```

### All temperature effects downscaling
If only concerned with temperature downscaling, the wrapper function `tempdownscale` allows all relevant processes to be run as a single function. 
```{r temp-dscale}
tmin<-tempdownscale( ukcp18rcm,
                  sst,# can be coarse resolution
                  dtmf, # fine scale dem
                  dtmm , # medium re wider are dem
                  basins = basins, # basindelin() output or will calculate
                  u2 = windspeed, # windspeeds downscaled - or will calculate
                  cad = TRUE,
                  coastal = TRUE,
                  tempvar = 'tmin',
                  whgto = 2, # output wind height above ground
                  thgto = 2 # output temperature height above ground
                  )

tmax<-tempdownscale(ukcp18rcm,
                  sst,# can be coarse
                  dtmf, # fine scale dem
                  dtmm , # medium re wider are dem
                  basins = basins, # basindelin() output
                  u2 = windspeed, # windspeeds downscaled
                  cad = TRUE,
                  coastal = TRUE,
                  tempvar = 'tmax',
                  whgto = 2, # output wind height above ground
                  thgto = 2 # output temperature height above ground
                  )

plot_q_layers(tmin)
plot_q_layers(tmax)

diurnaltmp<-tmax-tmin
names(diurnaltmp)<-rep('Diurnal_temp_range',nlyr(diurnaltmp))
plot_q_layers(diurnaltmp)

```

### Precipitation
The `precipdownscale()` function provides various options and methods to downscale rainfall. 


```{r precip-dscale}
method<-'Elev'
precelev<-precipdownscale(
  ukcp18rcm$climarray$prec,
  dtmf,
  ukcp18rcm$dtmc,
  method = method,
  fast = TRUE,
  noraincut = 0.01,
  patchsim = FALSE,
  nsim = dim( ukcp18rcm$climarray$prec)[3]
)
names(precelev)<-rep('prec_elev',nlyr(precelev))
terra::time(precelev)<-ukcp18rcm$tme
plot_q_layers(precelev)

method<-'Tps'
prectps<-precipdownscale(
  ukcp18rcm$climarray$prec,
  dtmf,
  ukcp18rcm$dtmc,
  method = method,
  fast = TRUE,
  noraincut = 0.01,
  patchsim = FALSE,
  nsim = dim( ukcp18rcm$climarray$prec)[3]
)
names(prectps)<-rep('prec_tps',nlyr(prectps))
terra::time(prectps)<-ukcp18rcm$tme

plot_q_layers(prectps)

prec<-prectps
```

### Longwave radiation downscale
Longwave downscaling accounts for the effects of terrain shading effect on skyview.
```{r lw-dscale}
# simple downscaling
lwf<-.resample(.rast(ukcp18rcm$climarray$lwrad,ukcp18rcm$dtmc), dtmf, msk=TRUE)
# corrected for skyview  
svf<-.rta(.skyview(dtmf),dim(lwf)[3])
lwrad<-.rast(.is(lwf)*svf,dtmf)

names(lwrad)<-rep('lwrad',nlyr(lwrad))
terra::time(lwrad)<-ukcp18rcm$tme
plot_q_layers(lwrad)

```

### Shortwave radiation downscale
Can be carried out without cloud patchiness or terrain shading...
```{r sw-dscale}
swrad<-swdownscale(ukcp18rcm$climarray$swrad,as.POSIXlt(ukcp18rcm$tme, tz = "UTC"),dtmf,ukcp18rcm$dtmc,patchsim = FALSE,terrainshade = FALSE)

names(swrad)<-rep('swrad',nlyr(swrad))
terra::time(swrad)<-ukcp18rcm$tme
                       
plot_q_layers(swrad)
```

... or with cloud patchiness...
```{r sw-patchdscale}
swradp<-swdownscale(ukcp18rcm$climarray$swrad,ukcp18rcm$tme,dtmf,ukcp18rcm$dtmc,patchsim = TRUE,nsim= length(ukcp18rcm$tme),terrainshade = FALSE)

names(swradp)<-rep('swrad_with_patchiness',nlyr(swradp))
terra::time(swradp)<-ukcp18rcm$tme
                       
plot_q_layers(swradp)
```
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
