---
title: "1. Data pre-processing"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# To run in developer mode: devtools::build_rmd('vignettes/articles/mesoclim_addtrees.Rmd')
```

```{r setup, include=FALSE}
library(lubridate)
library(terra)
#library(mesoclim)
devtools::load_all()
```

# 2 Preprocessing coarse resolution climate data

UKCP18 climate files are LARGE!! Downloadable from the CEDA archive they cover 10 years of data for the whole of the UK (Europe or world depending on the domain), each variable file is therefore typically > 100MB. Therefore it is important to define the area of interest and timespan for which downsclaed data is required.

## Defining area of interest

Spatial downscaling requires data covering not only for the immediate area of interest for which we require downscaled outputs, but also a wider area to ensure effects such as wind and coastal downscaling effects are captured.  

Digital elevation maps are required covering both the immediate and wider areas of interest. `mesoclim` makes use of three different extents and resolutions of dtm to inform its spatial downscaling:

* `dtmf` - a fine resolution digital terrrain map at covering the area and resolution at which downsclaed climate outputs are required.

* `dtmm` - a medium resolution digital terrain map used to calculate coastal effects.

* `dtmc` - a coarse resolution digital terrain map at the same resolution and projection as the original climate data. Typically this should be sourced from the providers of climate data so it matches the information used in climate modelling.The area covered should be the same of greater than that of the `dtmm`

All DTMs are assumed to cover land ONLY. Sea cells for all DTMs are expected to have the value NA. 

```{r inputs }
# Directory where downloaded .nc UKCP18 climate files are held
dir_ukcpdata<-"/Users/jonathanmosedale/Data/mesoclim_inputs"

# dir holding sea temperature data
dir_sst<-system.file('extdata/sst',package='mesoclim')

# Load coarse resolution DTM of the whole of the UK - provided by UKCP18
ukdtmc<-terra::rast(system.file('extdata/ukcp18rcm/orog_land-rcm_uk_12km_osgb.nc',package='mesoclim'))

# Define coarse resolution DTM of the widest area of interest from ukcp18 dtm
dtmc<-terra::crop(ukdtmc,ext(24000, 336000, -60000, 204000))

# Load medium resolution DTM (including wider area for capturing wind/coastal effects)
dtmm<-terra::rast(system.file('extdata/dtms/dtmm.tif',package='mesoclim'))

# Load a fine resolution dtm of the area and resolution for which we want downscaled climate values  
dtmf<-terra::rast(system.file('extdata/dtms/dtmf.tif',package='mesoclim'))

# Show local downscale area within wider dtmm
aoif<-vect(ext(dtmf))
aoim<-vect(ext(dtmm))
plot.new()
plot(dtmc,main='Medium and fine-resolution dtm extents on 12km coarse-resolution dtm')
plot(aoim, add=T)
plot(aoif,add=TRUE)
```

## Creation of standard inputs for downscaling

Different sources of climate and ancillary data can provide different variables using different SI units, in  varying formats and using different file naming conventions. Therefore, `mesoclim` provides several functions to convert and check common sources of climate data into standard inputs for subsequent downscaling. Here we look at UKCP18 regional data as an example.

### UKCP18 climate preprocessing

UKCP18 data pre-processing requires several modifications to the original data including:

* Conversion of 12 x 30-day months to actual dates.
* Conversion of net to downward short and longwave radiation. 

The conversion of shortwave radiation is calculated using an estimate of albedo at the same resolution as the climate data. Albedo data can either be provided or constant land/sea albedo values are used.

These conversions are carried out byt the function `ukcp18toclimarray`. Outputs are a list of climate variables and associated data which can be written to file using the function `write_climdata()`.

```{r preprocess, results="hide"}
# Preprocess UKCP18 data using constant albedo land / sea values
collection<-'land-rcm'
domain<-'uk'
member<-'01'
rcp<-'rcp85'
startdate<-as.POSIXlt('2018/05/01')
enddate<-as.POSIXlt('2018/05/31')

# Processes using already downloaded ukcp18rcm files in dir_data
t0<-now()
ukcp18rcm<-ukcp18toclimarray(dir_ukcpdata, dtmc,  startdate, enddate,collection, domain, member)
# should also work: ukcp18rcm<-ukcp18toclimarray(dir_ukcpdata, dtmm,  startdate, enddate,collection, domain, member)
print(now()-t0)

# Write preprocessed data
dir_out<-tempdir()
write_climdata(ukcp18rcm,file.path(dir_out,'ukcp18rcm.Rds'),overwrite=TRUE)
```

### UKCP18 sea surface temperature preprocessing

Sea surface temperature is not included in the UKCP18 data. For the UK seas surface data for the North Atlantic area for the same corresponding climate models used in the creation of the UKCP18 data is available on ceda. This data is in lat lon projection.

The function `create_ukcpsst_data` ensures that the sea surface data matches the corresponding dtmc in the UKCP18 climate data that has already been prepared - interpolating where necessary to ensure coast lines match. Sea surface data will match the crs, resolution and extent of the provided dtmc - which should extend well beyond the final downscaling area.

```{r sst-preprocess}
dir_sst<-system.file("extdata/sst", package = "mesoclim")
sst<-create_ukcpsst_data(dir_sst,startdate=as.POSIXlt('2018/05/01'),enddate=as.POSIXlt('2018/05/31'),dtmc=ukcp18rcm$dtm, member="01")
plot(c(sst,ukcp18rcm$dtm))
```

## Checking data inputs to downscaling

The resulting data structures of preprocessing can be checked to ensure there are no missing or unexpected values that may indicate a difference in the expected SI units or incomplete input datasets. This is particularly advisable if the inputs for spatial downscaling are not derived from one of the provided functions.

```{r check_inputs}
ukcp18rcm<-checkinputs(read_climdata(file.path(dir_out,'ukcp18rcm.Rds')), tstep = "day", plots=TRUE)
```
