---
title: "Mesoclimate downscaling on Jasmin"
---
Will use the ceda archive file structure to process UKCP18 regional climate and sea surface temperature data.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mesoclim)
```

# Setup & Inputs
All UKCP18 and HadUK gridded weather data is sourced from the ceda archive available on Jasmin.

## Downscaling requirements
Downscaling requires the following files & directories to be provided:

* Filepath to vector file of parcels output by ellicitor app.

* Filepath to coastline boundary polygon (not necessary if dtm already masked)

* Filepath to fine resolution DTM of UK (OS Terrain50 -  which has advantages in having a .vrt raster that can be queried to enhance cropping and extraction of relevant area to match aoi etc. Subdirectories within hold actual tiled data.)

* Directory for outputs - to which individual parcel .csv timeseries files are written.

Additional parameters required:

* Start time for future climate timeseries.

* End time for future climate timeseries.

* Model run of UKCP18rcm to be downscaled.

```{r downscale_inputs}
# CHANGE this...
dir_root<-'D:/'

# ... and these should be OK
parcels_file<-file.path(dir_root,'mesoclim_inputs','parcels','land_parcels.shp') # elicitor app output file
coast_file<-file.path(dir_root,'mesoclim_inputs','boundaries','CTRY_DEC_2023_UK_BGC.shp') # MHW line generalised to 20m
ukdtm_file<-file.path(dir_root,'mesoclim_inputs','dtm',"GBdem50.vrt") # 50m dtm virtual raster

dir_out<-file.path(dir_root,'mesoclim_inputs/outputs')  # output dir

# Parameters chosen to work with this data
ftr_sdate<-as.POSIXlt('2024/01/01')
ftr_edate<-as.POSIXlt('2039/12/31') # If using shared data folder use max value of as.POSIXlt('2039/12/31')

modelrun<-c('01') 
```

## Bias correction requirements
For bias correction the following additional directory and parameters are required: 

* Start time for historic bias correction calculations.

* End time for historic bias correction calculations.

* Directory holding UK ERA5 files for the bias correction time period.

```{r bias_inputs}
# CHANGE this...
dir_root<-'D:'

# ... and these should be OK
dir_era5<-file.path(dir_root,'era5')  # ERA5 data directory

# Timeperiod on which to carry out bias correction
pst_sdate<-as.POSIXlt('2016/01/01')
pst_edate<-as.POSIXlt('2021/12/31') 

modelrun<-c('01') 
```

# Prepare data 

## Area of interest and elevation data
Area of interest (aoi) for which downscaling will occur defined by parcel data and will be used for defining downscaling DTM and wider region for the cropping of climate inputs etc.

```{r aoi_dtms}
# Load UK fine resolution dtm & boundary(coast) 
dtm<-terra::rast(ukdtm_file) # 50m resolution dtm of all UK 
coast_v<-terra::project(terra::vect(coast_file),dtm)
  
# Load parcels file and project to crs of output dtm (OS coords)
parcels_v<-terra::project(terra::vect(parcels_file),dtm)

# Generate local and wider extents (for aoi and coatal/wind effects) 
aoi<-terra::vect(terra::ext(parcels_v))
crs(aoi)<-crs(dtm)

# Create fine resolution dtm of downscaling area  - ensure they fall within extent of loaded dtm & mask to coast_v (sets sea to NA)
dtmf<-terra::mask(terra::crop(terra::crop(dtm,aoi),dtm),coast_v)

# Plot dtmf and overlay parcels
plot(dtmf,main='Parcels')
plot(parcels_v,add=TRUE)
```



## Climate data for historic and future time periods


Climate and sea surface temperature data prepared from UKCP18 regional files on ceda archive

```{r ftrclimate_inputs}
t0<-now()
climdata<-addtrees_climarray(aoi,ftr_sdate,ftr_edate,collection='land-rcm',domain='uk',member='01')
sstdata<-addtrees_sstdata(ftr_sdate,ftr_edate,aoi=climdata$dtm,member='01')
prep_time= now()-t0

# write_climdata(climdata,file.path(dir_out,'ftr_ukcpdata.Rds'))
```

```{r histclimate_inputs}
t0<-now()
histdata<-addtrees_climarray(aoi,pst_sdate,pst_edate,collection='land-rcm',domain='uk',member='01')
prep_time= now()-t0
# write_climdata(climdata,file.path(dir_out,'ftr_ukcpdata.Rds'))
```


# Calculate bias correction models from historic data

## Observational data
For observed data we make use of ERA5 and HadUK 1km gridded data. Data is resampled to match UKCPrcm inputs CRS and resolution.

### ERA5

```{r era5_data}
yrs<-seq(year(pst_sdate),year(pst_edate),1)

# Get dtm informing era5 data
lsm<-rast(file.path(dir_era5,"lsm.tif"),subds='lsm')
geop<-rast(file.path(dir_era5,'geopotential.tif'),subds='z')
era5dtm<-mask(geop/9.80665,lsm,maskvalues=0)
plot(era5dtm)


# Load era5 yearly data
for(y in yrs){
  era5file<-file.path(dir_era5,paste0('era5_surface_ukeire_',y,'.nc'))
  file.exists(f)
  era5toclimarray(era5file, dtm=era5dtm, aoi=histdata$dtm, dtr_cor_fac = 1.285, toArrays=TRUE) 

}


dtmc<-terra::crop(terra::project(dtmc,crs(aoi)),aoi)
```


### HadUK termpareture and rainfall data

```{r get-haduk}
hadukdata<-addtrees_hadukdata(pst_sdate, pst_edate,climdata$dtm, varn=c('rainfall','tasmax','tasmin'),freq='day')

```


Model data could be directly downscaled to resolution of observed data (in this case 1km) but here we aggregate already downscaled data.

```{r get-moddata}
tmin<-.resample(unwrap(mesolizd$tmin),tminhuk[[1]])
tmax<-.resample(unwrap(mesolizd$tmax),tminhuk[[1]])
prec<-.resample(unwrap(mesolizd$prec),tminhuk[[1]])

```


## Calculate bias correction models for each variable

### Min daily temperature
```{r bias-tmincorrect}
tmn_bmods<-biascorrect(tminhuk, tmin, tmin,mod_out = TRUE, rangelims = NA)
print(tmn_bmods$models)
mn_stats<-c(app(tminhuk/tmin,mean),app(tminhuk/tmin,sd),app(tminhuk/tmin,max))
cf_mins<- c(app(tminhuk,min),app(tmin,min)) 
names(cf_mins)<-c('Yr tmn HadUK Observed','Yr tmn UKCP18 downsscaled')
plot(cf_mins)
```







# Downscaling future climate data

## Wider land-sea elevation data
A medium resolution DTM of the wider area is also prepared for use in windshelter and coastal corrections
```{r show_dsareas}
# Create medium resolution DTM for wider region around aoi
dtmm_res<-round(exp( ( log(terra::res(climdata$dtm)[1]) + log(terra::res(dtmf)[1]) ) / 2 ))

dtmm_f<-terra::mask(terra::crop(terra::crop(dtm,climdata$dtm),dtm),coast_v)
dtmm<-terra::mask(terra::aggregate(dtmm_f,dtmm_res / res(dtmf),  na.rm=TRUE),coast_v, touches=FALSE)

# Show local aoi within wider dtm
plot(dtmm,main='AOI area within wider dtmm')
plot(aoi,add=TRUE)
```







```{r check_inputs}
climdata<-checkinputs(climdata, tstep = "day")
```



