---
title: "ADDTREES Workflow without bias correction"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# To run in developer mode: devtools::build_rmd('vignettes/articles/mesoclim_addtrees.Rmd')
```

```{r setup}
library(lubridate)
library(terra)
library(mesoclim)
#devtools::load_all(quiet=TRUE)

tbegin<-now()
```


# Setup & Inputs
Following required files & directories:

* Filepath to vector file of parcels output by ellicitor app.

* Filepath to coastline boundary polygon (not necessary if dtm already masked)

* Filepath to 50m DTM of UK (OS Terrain50 -  which has advantages in having a .vrt raster that can be queried to enhance cropping and extraction of relevant area to match aoi etc. Subdirectories within hold actual tiled data.)

* Directory holding UKCP18 RCM climate data files - holding raw .nc files (by variable and decade) as downloaded from ceda. Must include files for historic UKCP18 rcm data for use in bias correction, as well as future files.

* Directory holding UKCP18 derived sea surface temperature data files - holding .nc files as downloaded from ceda

* Directory for outputs of parcel climate timeseries - to which individual parcel .csv files are written.

Additional parameters required:

* Start time for climate timeseries
* End time for climate series
* Model runs of UKCP18rcm to be analysed


For this article, I am assuming you are using the shared folder of mesoclim inputs. You should only have to change dir_root to the directory in which the shared folder is located for the rest of the setup to work...

For a quicker run time, reduce the time period to a single year.

```{r inputs}
# CHANGE this...
dir_root<-'D:'

# ... and these should be OK
parcels_file<-file.path(dir_root,'mesoclim_inputs','parcels','land_parcels.shp')
coast_file<-file.path(dir_root,'mesoclim_inputs','boundaries','CTRY_DEC_2023_UK_BGC.shp') # MHW line generalised to 20m
ukdtm_file<-file.path(dir_root,'mesoclim_inputs','dtm',"GBdem50.vrt") # 50m dtm virtual raster

dir_ukcp<-file.path(dir_root,'mesoclim_inputs/ukcp18rcm') # also holds ukcp rcm elevation data
dir_sst<-file.path(dir_root,'mesoclim_inputs/ukcp18sst')
dir_out<-file.path(dir_root,'mesoclim_inputs/outputs')

# Parameters chosen to work with this data
ftr_sdate<-as.POSIXlt('2020/01/01')
ftr_edate<-as.POSIXlt('2020/12/31') # If using shared data folder use max value of as.POSIXlt('2039/12/31')

modelrun<-c('06')

collection<-'land-rcm'
domain<-'uk'
rcp<-'rcp85'
```

# Download data 

*_ NOT required if working with the shared input data folder. _*

The functions will download from ceda archive the same data stored in the shared input data folder to the previously defined dir locations.

Download UKCP18 regional climate data: 

```{r download_climate, eval=FALSE}
# Download entire timeseries of observed and future UKCP data from ceda to dir_ukcp
report<-download_ukcp18(
    dir_ukcp,
    ftr_sdate,ftr_edate,
    collection, domain, rcp, member=modelrun,
    vars=c('clt','hurs','pr','prsn','psl','rls','rss','tasmax','tasmin','uas','vas'),
    download_dtm=TRUE, # will also download elevation data used in ukcp18 rcm model runs
    cedausr=cedausr, cedapwd=cedapwd
)
print(report)
```

Download UKCP18 driven sea surface temperature data:

```{r download_sst, eval=FALSE}
# sst data
report<-download_ukcpsst(dir_sst,ftr_sdate,ftr_edate,modelrun)
```


# Prepare  data
Area of interest (aoi) for which downscaling will occur defined by bounding box around parcel data. This is then used to crop elevation data to generate dtms required for downscaling.


```{r dtm_preprocess}
# Load dtm, boundary(coast) 
dtm<-terra::rast(ukdtm_file) # 50m resolution dtm of all UK 
coast_v<-terra::project(terra::vect(coast_file),dtm)
  
# Load parcels file and projectto crs of output dtmm (OS coords)
parcels_v<-terra::project(terra::vect(parcels_file),dtm)

# Generate local and wider extents (for aoi and coatal/wind effects) 
aoi<-terra::vect(terra::ext(terra::project(parcels_v,terra::crs(dtm))))
terra::crs(aoi)<-terra::crs(dtm)
buffer_metres<-units::set_units(15000,'m')
aoi_buf<-terra::buffer(aoi,buffer_metres)

# Create local dtms  - ensure they fall within extent of loaded dtm & mask to coast_v (sets sea to NA)
dtmf<-terra::mask(terra::crop(terra::crop(dtm,aoi),dtm),coast_v)
dtmm<-terra::mask(terra::crop(terra::crop(dtm,aoi_buf),dtm),coast_v)
dtmm<-terra::aggregate(dtmm,10)

# Show local aoi within wider dtm
plot(dtmm,main='AOI area within wider dtmm')
plot(aoi,add=TRUE)

# Coarse dtm matching ukcp18 inputs - crop to dtmm extent
dtmc<-terra::rast(file.path(dir_ukcp,"orog_land-rcm_uk_12km_osgb.nc"))
dtmc<-terra::crop(dtmc,dtmm) 

# Plot dtmf and overlay parcels
plot(dtmf,main='Parcels')
plot(parcels_v,add=TRUE)

# NOTE: might want to double check all parcels fall within land area of dtmf??
```

Future climate input files cropped and processed to a standard list of variables for the area of interest AND for each model run.
```{r prepare_climdata}
# Future
t0<-now()
ftr_ukcpdata<-ukcp18toclimarray(dir_ukcp, dtmc, ftr_sdate, ftr_edate,
                             collection, domain, modelrun)
print(now()-t0)

# write_climdata(ftr_ukcpdata,file.path(dir_out,'ftr_ukcpdata.Rds'))
```

Sea surface temperatures are also loaded and cropped to dtmm extent.
```{r sst_preprocess}
# Sea surface temperature 
sst<-create_ukcpsst_data(dir_sst,ftr_sdate,ftr_edate,dtmm,modelrun)
```

## Checking data inputs 
The resulting data structures of pre-processing can be checked to ensure there are no missing or unexpected values that may indicate a difference in the expected SI units or incomplete input datasets. This is particularly advisable if the inputs for spatial downscaling are not derived from one of the provided functions.Function also provides basic statistics and sample plots of climate variables.

```{r check_inputs}
ftr_ukcpdata<-checkinputs(ftr_ukcpdata, tstep = "day")
```


# Downscale ukcp18 climate data to 50m

```{r downscale_climate}
# Future - takes about 20mins using shared folder inputs
t0<-now()
ftr_mesoclim<-spatialdownscale(ftr_ukcpdata, sst, dtmf, dtmm, basins = NA, cad = TRUE,
                           coastal = TRUE, thgto = 2, whgto = 2, 
                           rhmin = 20, pksealevel = TRUE, patchsim = FALSE, 
                           terrainshade = FALSE, precipmethod = "Tps", fast = TRUE, noraincut = 0.01)
print(now()-t0)
```

### Display outputs

In tabular form for all times and locations:

```{r output_table}
climvars<-c('tmin','tmax','relhum','pres','swrad','lwrad','windspeed','winddir','prec')

smry_fun<-function(x) summary(as.vector(.is(x)))
rslt<-sapply(ftr_mesoclim[climvars],smry_fun)
stats_df<-as.data.frame(t(round(rslt,3)))[,c('Min.','Mean','Max.')]
print(stats_df)
```

Or as selected rasters for days corresponding to spatial quantiles (ie days where spatial means are data min, median and max values of timeseries):

```{r map-outputs}
for(var in climvars){
  r<-ftr_mesoclim[[var]]
  names(r)<-rep(var,nlyr(r))
  plot_q_layers(r,vtext=var)
}

# Show spatial range in daily temperatures
diurnaltmp<-ftr_mesoclim$tmax-ftr_mesoclim$tmin
plot_q_layers(diurnaltmp,vtext='diurnalT')
```

Or as timeseries of spatial statistics (min, max and mean), for example by day of year (averaged over all years) or by year:

```{r tseries_output}
par(mar=c(1,1,1,1),cex.main=0.8, mgp=c(3,0.1,0))
layout(matrix(c(1,2,3,4,1,5,6,7,1,8,9,10),ncol=3),heights=c(1,3,3,3))

#par(mfrow=c(3,3))
plot.new()
text(0.5,0.5,"Spatial mean (green), max (red) and min (blue) by day of year",cex=1,font=1)

# By Day of year (mean across timeseries)
for(v in climvars){
  r<-unwrap(ftr_mesoclim[[v]])
  plot_timestats_r(r,v,idx='doy',lgd=FALSE)
}

# By year
for(v in climvars){
  r<-unwrap(ftr_mesoclim[[v]])
  plot_timestats_r(r,v,idx='years',lgd=FALSE)
}

```


# Write outputs
Calculate weighted means of climate variables for each parcel and write as .csv files.

```{r write_outputs}
# Calculate parcel values 
t0<-now()
parcel_list<-create_parcel_list(ftr_mesoclim,parcels_v,id='gid')

# Write parcel files
write_parcels(parcel_list, dir_out, overwrite='append')
print(now()-t0)

```

```{r time-taken}
print(paste('Time taken for full analysis = ',now()-tbegin,'minutes' ))
```


