---
title: "2. Data pre-processing"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# To run in developer mode: devtools::build_rmd('vignettes/articles/mesoclim_addtrees.Rmd')
```

```{r setup}
library(lubridate)
library(terra)
library(mesoclim)
```

# 2 Preprocessing coarse resolution climate data

## Defining area of interest
Unless we are wanting to analyse the whole area covered by downloaded climate data, it is best to define a broad area of interest to restrict data size and processing times. The area of interest is used to crop a coarse DTM of the same projection and resolution as the climate data to provide a template for data processing. 

```{r inputs }
# dir holding already downloaded climate 
dir_ukcp<-system.file('extdata/ukcp18rcm',package='mesoclim')

# dir holding sea temperature data
dir_sst<-system.file('extdata/ukcp18sst',package='mesoclim')

# Load coarse resolution UKCP18rcm DTM of UK
dtmc<-terra::rast(file.path(dir_ukcp,'orog_land-rcm_uk_12km_osgb.nc'))

# Load DTM of general downscale area (including wider area for capturingwind/caosatal effects)
dtmm<-terra::rast(system.file('extdata/dtms/dtmm.tif',package='mesoclim'))

# Crop coarse DTM to an area of interest
dtmc<-terra::crop(dtmc,dtmm)

plot(dtmc,main='DTM of climate data extent to be extracted')
```

## Creation of standard inputs for downscaling
Different sources of climate and ancillary data can provide different variables using different SI units, in  different formats and using different file naming conventions. Therefore, `mesoclim` provides several functions to convert and check source data into standard inputs for subsequent downscaling. Here we look at UKCP18 regional data as an example.

### UKCP18 climate preprocessing
UKCP18 data preprocessing requires the conversion of net to downward short and longwave radiation. The conversion of shortwave radiation is calculated using an estimate of albedo at the same resolution as the climate data. Albedo data can either be provided or when not, constant land/sea albedo values are used.

Outputs are a list of climate variables which can be written to file using the function `write_climdata()`.

```{r preprocess, results="hide"}
# Preprocess UKCP18 data using constant albedo land / sea values
collection<-'land-rcm'
domain<-'uk'
member<-'01'
rcp<-'rcp85'
startdate<-as.POSIXlt('2018/05/01')
enddate<-as.POSIXlt('2018/05/30')

# Processes using already downloaded ukcp18rcm files in dir_data
t0<-now()
ukcp18rcm<-ukcp18toclimarray(dir_ukcp, dtmc,  startdate, enddate,
                             collection, domain, member)
print(now()-t0)

# Write preprocessed data
dir_out<-tempdir()
write_climdata(ukcp18rcm,file.path(dir_out,'ukcp18rcm.Rds'),overwrite=TRUE)
```

### UKCP18 sea surface temperature preprocessing
```{r sst-preprocess}
ukcp18sst<-create_ukcpsst_data(dir_sst,startdate,enddate,dtmc,member)
plot(ukcp18sst, range=range(global(ukcp18sst,'range',na.rm=TRUE)))
```

## Checking data inputs to downscaling
The resulting data structures of preprocessing can be checked to ensure there are no missing or unexpected values that may indicate a difference in the expected SI units or incomplete input datasets. This is particularly advisable if the inputs for spatial downscaling are not derived from one of the provided functions.

```{r check_inputs}
ukcp18rcm<-checkinputs(read_climdata(file.path(dir_out,'ukcp18rcm.Rds')), tstep = "day")
```
