---
title: "Temporal Downscaling"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
---

```{r, echo=FALSE,include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup, echo=FALSE, include=FALSE}
library(lubridate)
library(terra)
#library(mesoclim)
library(devtools)
load_all()
```


## Introduction

When both spatial and temporal dosnccaling is required (for example from UKCP18 data to hourly data at a higher resolution), it is advisable to **perform temporal downscaling prior to spatial downscaling**. This will help ensure that temporal fluctuation in, for example, windspeed and direction is incorporated into the spatial downscaling of not only these variable but dependent variables such as temperature. 

In this example, we use UKCP18RCM outputs for May 2018. 

### Some plot functions

Some useful functions for plotting hourly (solid lines) and daily (dotted lines) timeseries and spatial variation at times corrresponding to min/mode/max derived from spatial means at hourly inervals.

```{r plot-functions}
plot_r_quantiles<-function(r,p=c(0.01,0.5,1),maintxt){
  lyrstat<-global(r,mean,na.rm=TRUE)
  qtls<-quantile(lyrstat[,1],prob=p)
  sel<-c()
  for(q in qtls) sel<-c(sel,which.min(abs(lyrstat[,1] - q)))
  out_r<-r[[sel]]
  names(out_r)<-paste(maintxt,terra::time(r[[sel]]))
  plot(out_r,main=names(out_r),font.main=1, cex.main=1, nc=length(p))
}

# Plot hourly timeseries for contrasting days: spatial mean,min,max
plot_timeseries<-function(hrvar,dyvar=NA,vartext="Variable"){
  if(!inherits(dyvar,"logical")) vartext<-paste(vartext,"- solid hourly, dotted daily") else vartext<-paste(vartext,"- hourly")
  hr_mean<-unlist(global(hrvar, "mean", na.rm=TRUE))
  hr_max<-unlist(global(hrvar, "max", na.rm=TRUE))
  hr_min<-unlist(global(hrvar, "min", na.rm=TRUE))
  plot_hr<-as.data.frame(cbind(tstep=as.numeric(time(hrvar)),mean=hr_mean,max=hr_max,min=hr_min))
  plot_hr<-plot_hr[order(plot_hr$tstep),]
  matplot(as_datetime(plot_hr$tstep), plot_hr[,2:4], type = "l", lty = 1,
          col = c("black", "red", "blue"), xlab = "Date", ylab = vartext, font.main = 1,
          tck = 0.02, cex.main=1, cex.axis=0.7, main = vartext, cex.main=1)
  legend("topright", legend = c("Spatial Mean", "Spatial Max", "Spatial Min"), cex=0.5,
         col = c("black", "red", "blue"),
         lty = 1)
    
  if(!inherits(dyvar,"logical")){
    dy_mean<-unlist(global(dyvar, "mean", na.rm=TRUE))
    dy_max<-unlist(global(dyvar, "max", na.rm=TRUE))
    dy_min<-unlist(global(dyvar, "min", na.rm=TRUE))
    plot_dy<-as.data.frame(cbind(tstep=as.numeric(time(dyvar))+43200,mean=dy_mean,max=dy_max,min=dy_min))
    plot_dy<-plot_dy[order(plot_dy$tstep),]
    matplot(as_datetime(plot_dy$tstep), plot_dy[,2:4], type = "l", lty = 3,col = c("black", "red", "blue"),add=TRUE)
  }

}

# Load daily climate data
climdaily<-read_climdata(mesoclim::ukcpinput)
plot(climdaily$dtm)
```

## Single step downscaling

PROBLEM: LW rad - currently large daily leaps - CHECK LW method!!!

```{r singlestep_tempdscale}

allhrly<-temporaldownscale(climdaily, adjust = TRUE, clearsky=NA, srte = 0.09, relmin = 10, noraincut = 0)

for(n in 5:length(allhrly)) plot_timeseries(allhrly[[n]],dyvar=NA,vartext=names(allhrly)[n])
```

## Multi-step downscaling
As with spatial downscaling, the order in which variables are temporaly downscaled is important due to dependencies between the climate variables - for example the downscaling of daily humidity and longwave radiation to hourly requires the prior downscaling of temperature data.

### Daily to hourly temperature downscaling

Daytime temperatures are assumed to follow a sine curve with a peak a short while after solar noon. After dusk, the temperatures are assumed to decay exponentially reaching a minimum at dawn. The day in which tmx and tmn fall is assumed to match UTC days. The parameter `stre` controls the speed of decay of night time temperatures with time. A value of zero ensures values drop to minimum at dawn the following day, but trial and error indicates in most circumstances temperatures decay faster than this. The default value of 0.09 is an optimal value derived using ERA5 data for western Europe, but performs reasonably well globally

```{r temp_dscale}
# Generate hourly timeseries
hrtemps<-temp_dailytohourly(climdaily$tmin, climdaily$tmax, srte = 0.09) 

# Plot hourly timeseries for contrasting days: spatial mean,min,max
# Crude dilay tmean calculated for plotting comparison
dailymean<-(climdaily$tmax+climdaily$tmin)/2

# Plot rasters corresponding to 0.01, 0.5 and max quantiles (calculated fro spatial averages)
plot_r_quantiles(dailymean,p=c(0.01,0.5,1),maintxt="Daily mean temperature")
plot_r_quantiles(hrtemps,p=c(0.01,0.5,1),maintxt="Temperature")
plot_timeseries(hrtemps,dailymean,vartext="Temperature")

```

### Atmospheric pressure downscaling

Surface level pressures in kPa downscaled to hourly intervals. TheFunctions provided for converting between atmospheric and sea level pressure. 

```{r pres-downscale}
hrpres<-pres_dailytohourly(pres=climdaily$pres, tme=climdaily$tme, adjust = TRUE)

# Convert to sea level
hrpsl<-atmos_to_sea_pressure(hrpres,climdaily$dtm)

# Plots
plot_r_quantiles(hrpres,p=c(0.01,0.5,1),maintxt="Daily Atmos Pressure")
plot_r_quantiles(hrpres,p=c(0.01,0.5,1),maintxt="Hourly Atmos Pressure")
plot_timeseries(hrpres,climdaily$pres,vartext="Atmospheric Pressure")
```

### Relative humidity downscaling

Owing to the strong dependence of relative humidity on diurnal temperature patterns, prior to interpolation, relative humidity is first converted to specific humidity using tasmin, tasmax and psl. After interpolation, the data are back-converted to relative humidity, using temph and presh.

Function should spline interpolate vapour pressure and use temperature cycle

```{r humidity-downscale}
psl<-atmos_to_sea_pressure(climdaily$pres,climdaily$dtm)
hrrh<-hum_dailytohourly(relhum=climdaily$relhum, tasmin=climdaily$tmin, tasmax=climdaily$tmax, temph=hrtemps, psl=psl, presh=hrpres, tme=climdaily$tme, relmin = 2, adjust = TRUE)

# Plots
plot_r_quantiles(climdaily$relhum,p=c(0.01,0.5,1),maintxt="Daily Relative humidity")
plot_r_quantiles(hrrh,p=c(0.01,0.5,1),maintxt="Hourly Relative humidity")
plot_timeseries(hrrh,climdaily$relhum,vartext="Relative humidity")

```


### Shortwave radiation downscaling

Converting from daily to hourly downward shortwave radiation values is the most resource demanding of climate variables to downscale. 

Requires calculating the hourly and daily clearsky fraction (0:1) of radiation from which hourly clear sky radiation is calculated.

**Note:** The function assumes input radiation is downward flux, not net radiation (as provided in UKCP).

To get from net to downward flux we need to recognise that rswnet = (1-alb)*radsw, so radsw = rswnet/(1-alb), where alb is white sky albedo - as is carried out in the spatial downsclaing functions. However, white-sky albedo changes as a function of solar angle in a manner dependent on ground reflectance, leaf area, leaf inclination angles and leaf transmittance and the ratio of diffuse and direct radiation. The existing pre-processing of LW radiation data and temporal downscaling functions do NOT account for this hourly variation in albedo. 

These discrepancies are probably quite minor expect in areas with very low cover and should be largely captured by applying bias correction functions.

```{r swdaytohr}
swradhr<-swrad_dailytohourly(radsw=climdaily$swrad,  adjust = TRUE) 

plot_r_quantiles(climdaily$swrad,p=c(0.01,0.5,1),maintxt="Daily SW radiation")
plot_r_quantiles(swradhr,p=c(0.01,0.5,1),maintxt="Hourly SW radiation")
plot_timeseries(swradhr,climdaily$swrad, vartext="SW radiation")

```


### Downward longwave radiation downscaling

Currently this function recalculates LW down  from estimates of LW up from surface temperature and sky emissivity (??!!) before adjusting for daily LW radiation values:

(i) Calculate LW up using air temperature as approximation of surface temperature ie LWup =0.97*5.67*10**-8*(tc+273.15)

(ii) Calculates sky emissivity using dewpoint temperature (calculated from Tair, pressure and relative humidity) where skyem = 0.787 + 0.764 * log((tdp+273.15) / 273.15) - from Clark & Allan ref.

(iii) Calculates LW down as  Lwd = skyem * LWup 

(iv) If `adjust=TRUE`, adjusts daily average of hourly values reflect those of daily LW values.

If `adjust=FALSE`, the LW down inputs have no inluence but rather outputs will simply reflect temperature and calculated dew point temperature.

**Another option is to downscale from LWnet by simply removing temperture dependent LWup using hourly temps????**

```{r lw-downscale}
lwdhr<-lw_dailytohourly(lw=climdaily$lwrad, hrtemps=hrtemps, hrrh=hrrh, hrpres=hrpres, adjust = TRUE) 

# Compare spatraster daily/hourly quantiles
plot_r_quantiles(climdaily$lwrad,p=c(0.01,0.5,1),maintxt="LW radiation - daily")
plot_r_quantiles(lwdhr,p=c(0.01,0.5,1),maintxt="LW downward radiation - hourly")
plot_timeseries(lwdhr,climdaily$lwrad,vartext="LW radiation ")
```

### Wind speed downscaling

For interpolation, u and v wind vectors are derived form wind speed andd direction and these are interpolated to hourly, with backward calculations then performed to derive wind speed and direction.

* Need to spline interpolate u and v wind vectors. We could simulate inter-hourly variability. Follows a Weiball distribution so quite easy I suspect.

```{r wind-downscale}
hrwind<-wind_dailytohourly(climdaily$windspeed, climdaily$winddir, climdaily$tme, adjust = TRUE)

plot_r_quantiles(climdaily$windspeed,maintxt="Daily windspeed")
plot_r_quantiles(hrwind$wsh,maintxt="Daily windspeed")
plot_timeseries(hrwind$wsh,climdaily$windspeed, vartext="Windspeed")
```

### Precipitation downscaling

UNDER DEVELOPMENT!!

```{r precip-downscale}

# subdailyrain(rain=climdaily$prec, BLest, dailyvals = 24, dlim = 0.2, maxiter = 1000, splitthreshold = 0.2, trace = TRUE)


```
