---
title: "4. Temporal downscaling"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(lubridate)
library(terra)
library(mesoclim)
```

# 4. Temporal downscaling
Any efficiency or theoretical reasons for spatial before temporal downscaling or v.v?

```{r input-data}
# Get spatially downscaled data
 mesoclimate<-read_climdata(file.path(system.file('extdata/dscaledata',package='mesoclim'),'mesoclimate.rds'))
```

## Daily to hourly temperature downscaling
From one year's daily timeseries of spatially downscaled data.

```{r temp_tmedscale}

# Generate hourly timeseries
t0<-now()
hrtemps<-hourlytemp(mesoclimate$tmin, mesoclimate$tmax, srte = 0.09) 
#t<-as.POSIXlt(time(hrtemps))
print(now()-t0)

```

```{r graph-temp}
# Plot hourly timeseries for contrasting days: spatial mean,min,max
par(mfrow=c(1,2))
plot_timestats_r(hrtemps[[1:96]],v='Temperatures for 1-4 May',idx='hour',lgd=FALSE)

# Plot hourly rasters 
plot_q_layers(hrtemps,p=c(0, 0.5, 1),fun='mean', common_scale=FALSE)
```


Functions also permit daily to hourly downscaling of other climate variables.

Change variable names - tme parameter etc to standardise across functions

## Shortwave radiation downscaling
NOT working
```{r swdaytohr, eval=FALSE}
swradhr<-swrad_dailytohourly(mesoclimate$swrad, as.POSIXlt(terra::time(mesoclimate$swrad)), clearsky = NA, r = dtmf, adjust = TRUE) 

```
