---
title: "4. Bias correction" 
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
#library(mesoclim)
devtools::load_all(quiet=TRUE)

# Directories
dir_haduk1km<-"/Users/jonathanmosedale/Library/CloudStorage/OneDrive-UniversityofExeter/Data/haduk_1km_monthly"
dir_haduk12km<-"/Users/jonathanmosedale/Data/haduk_12km"
dir_terrain50<-"/Users/jonathanmosedale/Library/CloudStorage/OneDrive-UniversityofExeter/Data/Terrain50"
dir_ukcp<-"/Users/jonathanmosedale/Data/mesoclim_inputs"
dir_root<-"/Users/jonathanmosedale/Data"

# AOI etc
ukcpdtm<-rast(file.path(dir_root,"mesoclim_inputs","orog_land-rcm_uk_12km_osgb.nc"))
crs(ukcpdtm)<-"EPSG:27700"
#dtmf<-rast(system.file("extdata/dtms/dtmf.tif",package="mesoclim")) # already masked
#dtmm<-rast(system.file("extdata/dtms/dtmm.tif",package="mesoclim"))
#dtmc<-crop(ukcpdtm,dtmm)

dir_sst<-file.path(dir_root,"mesoclim_inputs")

```


# Functions 

Functions adapted  from mesoclimAddTrees to deal with HadUK data

```{r functions}
import_hadukdata<-function (startdate, enddate, dtmc=NA, filepath, haduk_res=c("1km","12km"),
    var = c("rainfall", "tasmax", "tasmin")) 
{
    varn <- match.arg(var)
    if (any(!var %in% c("rainfall", "tasmax", "tasmin"))) 
        stop("Chosen variables are not available as daily data!!")
    if (class(startdate)[1] != "POSIXlt" | class(enddate)[1] != 
        "POSIXlt") 
        stop("Date parameters NOT POSIXlt class!!")
    dateseq <- seq(as.Date(startdate) + 1, as.Date(enddate) + 
        1, by = "1 months") - 1
    yrs <- year(dateseq)
    mnths <- month(dateseq)
    mtxt <- ifelse(mnths < 10, paste0("0", mnths), paste0("", 
        mnths))
    daysofmonth <- c(31, 28, 31, 30, 31, 30, 31, 31, 30, 31, 
        30, 31)
    mdays <- daysofmonth[mnths]
    mdays <- ifelse((yrs%%4 == 0 & mnths == 2), 29, mdays)
    files <- paste0(var, "_hadukgrid_uk_",haduk_res,"_day_", yrs, mtxt, 
        "01-", yrs, mtxt, mdays, ".nc")
    filesin<-file.path(filepath,files)
    if(!all(file.exists(filesin))) stop("File for import_hadukdata not found!!!")
    r<-rast(filesin)
    crs(r)<-"EPSG:27700"
    if(!inherits(dtmc,"logical")){ 
      if (!terra::same.crs(dtmc, r))  dtmc <- terra::project(dtmc, r)
      e <- terra::crop(terra::extend(dtmc, 1), dtmc)
      #e <- terra::crop(terra::extend(terra::crop(r[[1]], dtmc, snap = "out"), 1), r[[1]])
      r <- terra::crop(r, e)
    }
    return(r)
}
```

# Introduction

Compares modeled historic data with observational data for the same period to either directly correct modelled data or provide bias-correction models for correction of modelled data.

Requires comparable raster stacks of observed and modeled data.

# Bias correction of temperature and precipitation
Firstly direct comparison of ukcp18rcm 12km data for 2018 and HadUK 12km gridded data from weather station observations. Grid is identical for these two sources of data. Haduk 12km data is only available for the same land/sea mask as UKCP18rcm. Yet many of the 'sea' cells not covered by HadUK 12km also contain land cells.

A landsea mask derived from Terrain 50m was caluclated for the UKCP18 12km grid to provide a proportion of land in each cell.

Derived climate data from 1km HadUK - aggregate to 12km grid - compare with existing 12km 

## Land / sea proportion in UKCP data

No land/sea proportions available for UKCP18rcm data so calculated using Terrain50 and coastal mask.

* All UK boundary data typically excludes Isle of Ma (and land sea mask can be created from lower res data).
* OS Terrain50 excludes Northern Ireland

Therefore only consider coatal cells for GB excluding Isle of Man and NI.

See `create_ukcp18rcm_lsm.R` for code used in creation of 12km land sea proportions.

```{r get_landsea_mask)}
# Proportion land 12km derived from Terrain50 - GB only!!
lsm12km.r<-rast("/Users/jonathanmosedale/Library/CloudStorage/OneDrive-UniversityofExeter/Data/Terrain50/uk_seamask_12km.tif")
plot(lsm12km.r)
```

## UKCP input data 

Load UKCP data matching 12km landsea mask

```{r ukcpdata}
dtmc<-crop(ukcpdtm,lsm12km.r)
plot(dtmc)
collection<-'land-rcm'
domain<-'uk'
member<-'01'
rcp<-'rcp85'
startdate<-as.POSIXlt('2018/01/01') 
enddate<-as.POSIXlt('2018/12/31')

climdata<-ukcp18toclimarray(dir_ukcp,dtmc,startdate,enddate,collection,domain,member,temp_hgt=1.5, wind_hgt=10,toArrays=FALSE)


```

# HadUK 12km data

```{r cf_obs_ukcp_12km}
# Get daily Haduk data for UK@ 12km 
tasmax12km.r<-import_hadukdata(startdate, enddate, dtmc=lsm12km.r, filepath=dir_haduk12km, haduk_res="12km",
    var = "tasmax") 
tasmin12km.r<-import_hadukdata(startdate, enddate, dtmc=lsm12km.r, filepath=dir_haduk12km, haduk_res="12km",
    var = "tasmin") 
#rain12km.r<-import_hadukdata(startdate, enddate, dtmc, filepath=dir_haduk12km, haduk_res="12km", var = "rainfall") 

# Mask using ls12km land only - removes Isle of Man and Northern Ireland
tasmax12km.r<-mask(tasmax12km.r,lsm12km.r,maskvalue=0)
tasmin12km.r<-mask(tasmin12km.r,lsm12km.r,maskvalue=0)
#rain12km.r<-mask(rain12km.r,lsm12km.r,maskvalue=0)
diurnal12km.r<-tasmax12km.r-tasmin12km.r

plot(c(tasmax12km.r[[1]],climdata$tmax[[1]]))
```

## Observational data 1km

For observed data we make use of the HadUK 1km gridded data for the Lizard during the year 2018.

```{r get-obsdata}

tmin1km<-import_hadukdata(startdate, enddate, NA, filepath=dir_haduk1km, haduk_res="1km", var = "tasmin") 
tmax1km<-import_hadukdata(startdate, enddate, NA, filepath=dir_haduk1km, haduk_res="1km", var = "tasmax") 
#rain1km<-rast(system.file("extdata/haduk/rainfall1km.tif",package="mesoclim"))

# Extend/Crop to align with 12km grid then aggregateResample to 12km res using average
tmin1km<-crop(extend(tmin1km,ext(tasmin12km.r)),ext(tasmin12km.r))
tmin12kmb<-terra::aggregate(tmin1km,12,fun="mean",na.rm=TRUE)

tmax1km<-crop(extend(tmax1km,ext(tasmax12km.r)),ext(tasmax12km.r))
tmax12kmb<-terra::aggregate(tmax1km,12,fun="mean",na.rm=TRUE)
#rain1km<-crop(extend(rain1km[[1]],ext(rain12km.r)),ext(rain12km.r))
#rain12kmb<-terra::aggregate(rain1km,12,fun="mean",na.rm=TRUE)

# mask out all sea cells (and NI and IoM)
tmin12kmb<-mask(tmin12kmb,lsm12km.r,maskvalue=0)
tmax12kmb<-mask(tmax12kmb,lsm12km.r,maskvalue=0)
#rain12kmb<-mask(rain12kmb,lsm12km.r,maskvalue=0)

diurnal12kmb<-tmax12kmb-tmin12kmb

plot(c(tmax12kmb[[5]],tmin12kmb[[5]],diurnal12kmb[[5]],lsm12km.r))
```

### Compare  HAdUK 12km gridded data with aggregated 1km data 

This only compares cells where ukcp18rcm data - ignoring the many coastal regions not covered by the original 1km data
Very little difference between 12km downloaded data and that derived from agregation of 1km data:

* Tmax - small cluster in NW Scotlang where 12km cells are up to 1.5C LOWER than aggregated 1km
* Tmin - small cluster  in borders where 12km cells are up to 0.9C GREATER than aggregated 1km - likely error in 12km data

```{r cf_1km_12km_HadUK}
tmin_anom<- tasmin12km.r-mask(tmin12kmb,tasmin12km.r)
tmax_anom<- tasmax12km.r-mask(tmax12kmb,tasmax12km.r)
tdnl_anom<- diurnal12km.r-mask(diurnal12kmb,diurnal12km.r)

par(mfrow=c(1,3))
plot(c(mean(tmin_anom),mean(tmax_anom),mean(tdnl_anom)),main="Mean anomaly")
plot(c(max(tmin_anom),max(tmax_anom),max(tdnl_anom)),main="Max anomaly")
plot(c(min(tmin_anom),min(tmax_anom),min(tdnl_anom)),main="Min anomaly")

```

## Compare Haduk data at 12km with UKCP18 data at 12km



```{r cf_1km_12km_HadUK}
tmin_anom<- mask(climdata$tmin,tasmin12km.r)-mask(tmin12kmb,tasmin12km.r)
tmax_anom<- climdata$tmax-mask(tmax12kmb,tasmax12km.r)
tdnl_anom<- (climdata$tmax-climdata$tmin)-mask(diurnal12kmb,diurnal12km.r)

par(mfrow=c(1,3))
plot(c(mean(tmin_anom),mean(tmax_anom),mean(tdnl_anom)),main="Mean anomaly")
plot(c(max(tmin_anom),max(tmax_anom),max(tdnl_anom)),main="Max anomaly")
plot(c(min(tmin_anom),min(tmax_anom),min(tdnl_anom)),main="Min anomaly")

```



## Bias correction of 12km model and observed data

Purpose: to ensure mean predicted/model values match mean observed values for the same historic time period.

Comparison and bias correction of all UKCP cells with 50m land cells within.

Carried out on raw 12km UKCP variables after initial processing but no downscaling.

Daily observed historic data sourced from HadUK (tmin,tmax, rainfall) and from temporally aggregated and resampled ERA5 data:






### Min daily temperature

Observed has much greater range (+/-10C) in with a mean anomaly -0.5 to -5 C

```{r bias-12kmtmincorrect}
# Set mask to either only UKCP land cells or where some land in lsm12km.r
#msk<-tasmin12km.r[[1]]
msk<-ifel(lsm12km.r==0,NA,lsm12km.r)

tmn_bmods<-biascorrect(hist_obs=mask(tmin12kmb,msk), hist_mod=mask(climdata$tmin,msk), fut_mod = NA,mod_out = TRUE, rangelims = NA)
print(tmn_bmods$models)

# Compare mean values
obs_mean<-app(mask(tmin12kmb,msk),mean)
mod_mean<-app(mask(climdata$tmin,msk),mean)
cf_means<- c(obs_mean,mod_mean,obs_mean-mod_mean)
names(cf_means)<-c('Observed','Modelled','Anomaly')
plot(cf_means)

# Compare min / max values
obs_min<-app(mask(tmin12kmb,msk),min)
mod_min<-app(mask(climdata$tmin,msk),min)
cf_mins<- c(obs_min,mod_min,obs_min-mod_min)
names(cf_mins)<-c('Observed','Modelled','Anomaly')
plot(cf_mins)

obs_max<-app(mask(tmin12kmb,msk),max)
mod_max<-app(mask(climdata$tmin,msk),max)
cf_max<- c(obs_max,mod_max,obs_max-mod_max)
names(cf_max)<-c('Observed','Modelled','Anomaly')
plot(cf_max)
```

```{r apply_tmin_bcmodel}
# Apply bias correction to data
tmin_bc<-biascorrect_apply(fut_mod=mask(climdata$tmin,msk), biasmods=tmn_bmods, rangelims = NA)

# Calculate bc anomalies
cf_mins<- c(app(mask(tmin12kmb,msk),min),app(mask(tmin12kmb,msk),max),app(tmin_bc,min),app(tmin_bc,max)) 
names(cf_mins)<-c('Observed min','Observed max','BC model min','BC model max')
plot(cf_mins)
dif<-mask(tmin12kmb,msk)-mask(tmin_bc,msk)
mn_stats<-c(app(dif,mean),app(dif,sd),(app(mask(tmin12kmb,msk),max)-app(mask(tmin12kmb,msk),min))-(app(tmin_bc,max)-app(tmin_bc,min)))
names(mn_stats)<-c("Mean obs-bc_model","sd obs-bc_model","obs range - bc_model range")
plot(mn_stats)
```

### Max daily temperature

Observed show much greater range than modelled (UKCP) (+/-8C) and a mean anomaly from 1.5 to -3.5 C

```{r bias-1kmtmaxcorrect}
tmx_bmods<-biascorrect(hist_obs=mask(tmax12kmb,msk), hist_mod=mask(climdata$tmax,msk), fut_mod = NA, mod_out = TRUE, rangelims = NA, samplenum=1000)
print(tmx_bmods$models)

# Compare mean values
obs_mean<-app(mask(tmax12kmb,msk),mean)
mod_mean<-app(mask(climdata$tmax,msk),mean)
cf_means<- c(obs_mean,mod_mean,obs_mean-mod_mean)
names(cf_means)<-c('Observed','Modelled','Anomaly')
plot(cf_means)

# Compare min / max values
obs_min<-app(mask(tmax12kmb,msk),min)
mod_min<-app(mask(climdata$tmax,msk),min)
cf_mins<- c(obs_min,mod_min,obs_min-mod_min)
names(cf_mins)<-c('Observed','Modelled','Anomaly')
plot(cf_mins)

obs_max<-app(mask(tmax12kmb,msk),max)
mod_max<-app(mask(climdata$tmax,msk),max)
cf_max<- c(obs_max,mod_max,obs_max-mod_max)
names(cf_max)<-c('Observed','Modelled','Anomaly')
plot(cf_max)
```

```{r apply_12kn_bc}
# Apply bias correction to data
tmax_bc<-biascorrect_apply(fut_mod=mask(climdata$tmax,msk), biasmods=tmx_bmods, rangelims = NA)

# Calculate bc anomalies
cf<- c(app(mask(tmax12kmb,msk),min),app(mask(tmax12kmb,msk),max),app(tmax_bc,min),app(tmax_bc,max)) 
names(cf)<-c('Observed min','Observed max','BC model min','BC model max')
plot(cf)
dif<-mask(tmax12kmb,msk)-mask(tmax_bc,msk)
stats<-c(app(dif,mean),app(dif,sd),(app(mask(tmax12kmb,msk),max)-app(mask(tmax12kmb,msk),min))-(app(tmax_bc,max)-app(tmax_bc,min)))
names(stats)<-c("Mean obs-bc_model","sd obs-bc_model","obs range - bc_model range")
plot(stats)
```

### Rainfall


## ERA 5 variables

### Convert to daily
Writes .tifs of daily values of each ERA5 variable.
.era5todaily(filein,pathout,landsea) 
Any other processing of variables required??

UKCP processing:

* Pressure: compare and correct psl (not atmos) 
* RH: 
* Windspeed: from u/v vectors convert to common height??
* SWrad - correct UKCP to downward (albedo from finescale lsmask?)
* LWrad - correct UKCP to downward

ERA5 processing

* All to be converted from hourly to daily means
* All to be resampled to UKCP resolution and grid
* Pressure: compare and correct psl (not atmos) 
* RH: 
* Windspeed: from u/v vectors convert to same height as UKCP??
* SWrad - correct UKCP to downward (albedo from finescale lsmask?)
* LWrad - correct UKCP to downward


```{r}

.era5todaily(filein,pathout,landsea, startdate,enddate)

```








## Historic modelled data

Create model outputs for the same historical period as the observational data. In this case we directly downscale to the same resolution and extent as the observed data (1km). Alternatively, if downscaled to a finer resolution outputs could be aggregated to observed data resolution

```{r input-data}
# Must unpack spatrasters etc saved in .rda data object
ukcpinput<-read_climdata(mesoclim::ukcpinput)
ukcp18sst<-rast(mesoclim::ukcp18sst)

# Get spatially downscaled data to 1km
dtmf<-rast(system.file("extdata/dtms/dtmf.tif",package="mesoclim")) %>% aggregate(20,na.rm=TRUE) %>% mask(tmin1km[[1]])

dtmm<-rast(system.file("extdata/dtms/dtmm.tif",package="mesoclim")) %>% aggregate(5,na.rm=TRUE)

plot(dtmm); plot(dtmf,add=T)

modeldata1km<-spatialdownscale(climdata=ukcpinput, sst=ukcp18sst, dtmf, dtmm, basins = NA,  wca=NA, cad=TRUE,
                     coastal = TRUE, thgto =2, whgto=2, rhmin = 20, pksealevel = TRUE, patchsim = FALSE, 
                     terrainshade = TRUE, precipmethod = "Elev", fast = TRUE, noraincut = 0.01, toArrays=FALSE)
```

## Calculate bias correction models for each variable

Calculate bias correction models for each climate variable. MORE

### Min daily temperature

```{r bias-tmincorrect}
tmn_bmods<-biascorrect(hist_obs=tmin1km, hist_mod=modeldata1km$tmin, fut_mod = NA,mod_out = TRUE, rangelims = NA)
print(tmn_bmods$models)

cf_mins<- c(app(tmin1km,min),app(modeldata1km$tmin,min)) 
names(cf_mins)<-c('Observed min(Tmin) of time period','Model min(Tmin) of time period')
plot(cf_mins)

dif<-tmin1km-modeldata1km$tmin
mn_stats<-c(app(dif,mean),app(dif,sd),app(dif,max),app(dif,min))
names(mn_stats)<-c("Mean obs-model","sd obs-model","Max obs-model","Min obs-model")
plot(mn_stats)

```

### Max daily temperature

```{r bias-tmaxcorrect}
tmx_bmods<-biascorrect(hist_obs=tmax1km, hist_mod=modeldata1km$tmax, fut_mod = NA, mod_out = TRUE, rangelims = NA, samplenum=1000)
print(tmx_bmods$models)
cf_maxs<- c(app(tmax1km,max),app(modeldata1km$tmax,max)) 

names(cf_maxs)<-c('Observed Tmax of time period','Model Tmax of time period')
plot(cf_maxs)

dif<-tmax1km-modeldata1km$tmax
mx_stats<-c(app(dif,mean),app(dif,sd),app(dif,max),app(dif,min))
names(mx_stats)<-c("Mean obs-model","sd obs-model","Max obs-model","Min obs-model")
plot(mx_stats)
```

### Daily precipitation

```{r bias-precipcorrect}
prec_bmods<-precipcorrect(prechuk, prec, prec, mod_out = TRUE, rangelim = NA)

plot(unwrap(prec_bmods$mu_tot),main='Total rainfall')
plot(unwrap(prec_bmods$mu_frac ),main='Proportion of rainy days')
```



# Bias correction of all variables


## Bias correct inputs and parameters

```{r bias_params}
# Files
era5landsea


# Directories
dir_era5
dir_era5daily

dir_ukcp
dir_ukcpera5




```


## Prepare ERA5 observational data


### Any correction of temp?





## Prepare UKCP18 climate data
Future climate input files cropped and processed. 

Does it require to a standard list of variables at this stage??


```{r prepare_climdata}
ukcptoera5<-function(dir_ukcp,era5landsea,startdate,enddate,
                collection=c('land-gcm','land-rcm'),
                domain=c('uk','eur','global'),
                member=c('01','02','03','04','05','06','07','08','09','10','11','12','13','14','15','16','17','18','19','20','21','22','23','24','25','26','27','28'),
                ukcp_vars=c('clt','hurs','pr','psl','rls','rss','tasmax','tasmin','uas','vas') ){
                  
  # Parameter check
  collection<-match.arg(collection)
  domain<-match.arg(domain)
  ukcp_vars<-match.arg(ukcp_vars,several.ok=TRUE)
  ukcp_units<-match.arg(ukcp_units,several.ok=TRUE)
  output_units<-match.arg(output_units,several.ok=TRUE)
  if(class(startdate)[1]!="POSIXlt" | class(enddate)[1]!="POSIXlt") stop("Date parameters NOT POSIXlt class!!")
  
  # Set sea cells to NA
  era5landsea[era5landsea==0]<-NA
  
  # Check member in chosen collection
  member<-match.arg(member)
  if(collection =='land-rcm' &
     !member %in% c('01','02','03','04','05','06','07','08','09','10','11','12','13','14','15')) stop(paste("Model member",member,"NOT available in land-rcm collection - ignoring!!"))

  # Check dir_ukcp exists
  if(!dir.exists(dir_ukcp)) stop(paste("Directory does NOT exist:",dir_ukcp))

  # Identify which decades are required
  decades<-.find_ukcp_decade(collection,startdate,enddate)

  # Get ukcp files names and check all exist in dir_ukcp before loading
  if(collection=='land-gcm') collres<-'60km' else collres<-'12km'
  ncfiles<-do.call(paste0, c(expand.grid(ukcp_vars,'_rcp85_',collection,'_',domain,'_',collres,'_',member,'_day_', decades,".nc")))
  ncfiles<-file.path(dir_ukcp,ncfiles)
  not_present<-which(!file.exists(ncfiles))
  if (length(not_present)>0) stop(paste("File NOT present in ukcp directory: ",ncfiles[not_present]," ") )

  # For each UKCP variable crop and res
  # Reproject and resample to landsea
  # apply landsea mask
  for (n in 1:length(ukcp_vars)){
    v<-ukcp_vars[n]
    var_r<-terra::rast()
    
    for (d in decades){
      filename<-paste0(v,'_rcp85_',collection,'_',domain,'_',collres,'_',member,'_day_', d,".nc")
      message(paste("Loading",filename))
      ncfile<-file.path(dir_ukcp, filename)
      
      # Load and resample data to ERA5 landsea
      r <- terra::rast(ncfile, subds=v)
      r<-.resample(r,landsea)
      # Mask sea 
      r<-terra::crop(r,dtmc)
      r<-terra::mask(r,era5landsea)
      
      # Convert to real calendar dates and fill missing dates
      ukcp_dates<-.get_ukcp18_dates(ncfile)
      real_dates<-.correct_ukcp_dates(ukcp_dates)
      r<-.fill_calendar_data(r, real_dates)
    }
    
    # Join if multiple decades of data and write tif
    terra::add(var_r)<-r
    writeRaster(var_r,file.path(pathout,filename))
    }
} # end function

```



## Checking data inputs 
The resulting data structures of pre-processing can be checked to ensure there are no missing or unexpected values that may indicate a difference in the expected SI units or incomplete input datasets. This is particularly advisable if the inputs for spatial downscaling are not derived from one of the provided functions.

The `checkinputs` function also provides basic statistics and sample plots of climate variables.

```{r check_inputs}
ftr_ukcpdata<-checkinputs(ftr_ukcpdata, tstep = "day")
```


# Bias correct ukcp18 climate data using ERA5 data

```{r ukcp_biascorrect}
# Future - takes about 20mins using shared folder inputs
t0<-now()

model_list<-list()
ukcp_vars<-c('tmin','tmax','relhum','pres','swrad','lwrad','windspeed','winddir','prec')
era5_vars
for (v in ukcp_vars)
bcmodel<-biascorrect(era5_obs, ukcp_mod, fut_mod = NA, mod_out = TRUE, rangelims = NA)
  

#biascorrectukcpall<-function(pathtoera,pathtoukcp18,pathtoukcpdecade,pathout,decade,modelrun) {


bc_time<-now()-t0
```

Time taken for downscaling function = `r round(bc_time,2)`  minutes.

### Display downscaled outputs

Min, max and mean values in tabular form for all times and locations:

```{r output_table}
climvars<-c('tmin','tmax','relhum','pres','swrad','lwrad','windspeed','winddir','prec')

smry_fun<-function(x) summary(as.vector(as.array(x)))
rslt<-sapply(ftr_mesoclim[climvars],smry_fun)
stats_df<-as.data.frame(t(round(rslt,3)))[,c('Min.','Mean','Max.')]
print(stats_df)
```

Or as raster plots for days corresponding to spatial quantiles (ie days where spatial means are data min, median and max values of timeseries). Example plots of max and min daily temperatures, precipitation and dirunal temperature range:

```{r map-outputs}
for(var in c('tmax','tmin','prec')){
  r<-ftr_mesoclim[[var]]
  names(r)<-rep(var,nlyr(r))
  plot_q_layers(r,vtext=var)
}

# Show spatial range in daily temperatures
diurnaltmp<-ftr_mesoclim$tmax-ftr_mesoclim$tmin
plot_q_layers(diurnaltmp,vtext='diurnalT')
```

Or as plots of timeseries of spatial statistics (min, max and mean), for example by day of year (averaged over all years):
  
```{r tseries_output}
par(mar=c(1,1,1,1),cex.main=0.8, mgp=c(3,0.1,0))
layout(matrix(c(1,2,3,4,1,5,6,7,1,8,9,10),ncol=3),heights=c(1,3,3,3))
plot.new()
text(0.5,0.5,"Spatial mean (green), max (red) and min (blue) by day of year",cex=1,font=1)
for(v in climvars){
  r<-unwrap(ftr_mesoclim[[v]])
  plot_timestats_r(r,v,idx='doy',lgd=FALSE)
}
```
