---
title: "5. Bias correction"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
#library(mesoclim)
devtools::load_all(quiet=TRUE)
```

# 5 Bias Correction
Compares modeled historic data with observational data.
Requires comparable raster stacks of observed and modeled data.

## Observational data
For observed data we make use of the HadUK 1km gridded data for the Lizard during the year 2018.
```{r get-obsdata}
dir_haduk<-system.file('extdata/haduk',package='mesoclim')

tminfiles<- list.files(dir_haduk, pattern='tasmin')
tmaxfiles<- list.files(dir_haduk, pattern='tasmax')
rainfiles<- list.files(dir_haduk, pattern='rainfall')

tminhuk<-rast(file.path(dir_haduk,tminfiles))
tmaxhuk<-rast(file.path(dir_haduk,tmaxfiles))
prechuk<-rast(file.path(dir_haduk,rainfiles))

# Crop to area of interest
dir_datasets<-system.file('data',package='mesoclim')
load(file.path(dir_datasets,'dtmf.rda'))
dtmf<-unwrap(dtmf)

e<-ext(project(dtmf,crs(tminhuk)))
tminhuk<-crop(tminhuk,e)
tmaxhuk<-crop(tmaxhuk,e)
prechuk<-crop(prechuk,e)

```

## Historic modelled data
Use model outputs for the same historical period as the observational data.
```{r input-data}
# Get spatially downscaled data
 mesolizd<-read_climdata(file.path(system.file('extdata/dscaledata',package='mesoclim'),'mesolizd.rds'))
```

Model data could be directly downscaled to resolution of observed data (in this case 1km) but here we aggregate already downscaled data.

```{r get-moddata}
tmin<-.resample(unwrap(mesolizd$tmin),tminhuk[[1]])
tmax<-.resample(unwrap(mesolizd$tmax),tminhuk[[1]])
prec<-.resample(unwrap(mesolizd$prec),tminhuk[[1]])

```


## Calculate bias correction models for each variable

### Min daily temperature
```{r bias-tmincorrect}
tmn_bmods<-biascorrect(tminhuk, tmin, tmin,mod_out = TRUE, rangelims = NA)
print(tmn_bmods$models)
mn_stats<-c(app(tminhuk/tmin,mean),app(tminhuk/tmin,sd),app(tminhuk/tmin,max))
cf_mins<- c(app(tminhuk,min),app(tmin,min)) 
names(cf_mins)<-c('Yr tmn HadUK Observed','Yr tmn UKCP18 downsscaled')
plot(cf_mins)
```

### Max daily temperature

```{r bias-tmaxcorrect}
tmx_bmods<-biascorrect(tmaxhuk, tmin, tmin,mod_out = TRUE, rangelims = NA)
print(tmx_bmods$models)
cf_maxs<- c(app(tmaxhuk,max),app(tmax,max)) 
names(cf_maxs)<-c('Yr tmx HadUK Observed','Yr tmx UKCP18 downsscaled')
plot(cf_maxs)

mx_stats<-c(app(abs(tmaxhuk/tmax),mean),app(abs(tmaxhuk/tmax),sd),app(abs(tmaxhuk/tmax),max))
plot(mx_stats)
```

### Daily precipitation

```{r bias-precipcorrect}
prec_bmods<-precipcorrect(prechuk, prec, prec, mod_out = TRUE, rangelim = NA)

plot(unwrap(prec_bmods$mu_tot),main='Total rainfall')
plot(unwrap(prec_bmods$mu_frac ),main='Proportion of rainy days')
```
